---
title: "kenya_code"
author: "Dianovics Dominik"
date: "2023-12-15"
output:
  html_document: default
  pdf_document: default
---

# Final project using Kenya census 2019

[Github tidytuesday link](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-01-19/readme.md)

```{r setup, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidytuesdayR)
library(tidyverse)
library(tools)
library(stringdist)
library(fuzzyjoin)
library(patchwork)
library(knitr)
library(gt)
library(grid)
library(gridExtra)
library(broom)
library(lmtest)
library(robustbase)
library(MASS)
library(mediation)
library(devtools)
library(scales)
library(flexplot)
library(rempsyc)
library(apaTables)

options(scipen = 999)
```

## Firstly, **import** the data

```{r data import, echo=TRUE, results='hide'}
data <- tt_load('2021-01-19')

gender <- data$gender
crops <- data$crops
households <- data$households
```

## Then data **cleaning**

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
view(crops)
str(crops)

view(households)
str(households)

view(gender)
str(gender)
```

```{r data cleaning}
#Issue with crops, all names are caps lock
crops <- crops |> 
  mutate(
    SubCounty = str_to_lower(SubCounty),
    SubCounty = str_to_title(SubCounty)
  )

#Removed the Khat crop because I don't know what that is and there are too many missing values
crops <- crops[, !colnames(crops) %in% "Khat (Miraa)"]


#Issue with city naming, not consistent through datasets
crops <- crops |> 
  mutate(
    County = SubCounty,
  ) |> 
  dplyr::select(-SubCounty)

#Changing the naming from Total to Kenya
gender <- gender |> 
  mutate(
    County = ifelse(County == "Total", "Kenya", County)
  )

#Cleaning whitespaces
gender$County <- gsub(" ", "", gender$County)
households$County <- trimws(households$County)
```

```{r matching function and merge}
#Function that tries to match the County names to each other by stringdist
find_best_match <- function(county, reference_counties) {
  distances <- stringdistmatrix(county, reference_counties)
  best_match_index <- which.min(distances)
  best_match <- reference_counties[best_match_index]
  return(best_match)
}

#Function had an issue with only one value, the Nairobi city, probably due to two words in the name
crops$County[crops$County == "Nairobi"] <- "Nairobi City"
crops$best_match <- sapply(crops$County, find_best_match, reference_counties = gender$County)
households$best_match <- sapply(households$County, find_best_match, reference_counties = gender$County)

#Merge datasets
merged_dataset <- left_join(crops, households, by = c("best_match" = "County")) |> 
  left_join(gender, by = c("best_match" = "County"))

# Clean up 
merged_dataset <- merged_dataset |> 
  dplyr::select(-best_match, -best_match.y)

#County in the first column
merged_dataset <- merged_dataset |> 
  dplyr::select(County, everything())
```

# Data exploration

## **Gender**

```{r}
#Gender in the whole of Kenya
data_kenya <- filter(merged_dataset, County == "Kenya")
data_kenya_long <- pivot_longer(data_kenya, cols = c("Male", "Female", "Intersex"), names_to = "Gender", values_to = "Count")
data_kenya_long |> 
  ggplot(aes(x = reorder(Gender, -Count), y = Count)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_number_si()) +
  coord_flip() +
  theme_classic() +
  labs(
    y = "Count",
    x = "Gender"
  ) +
  ggtitle("Gender distribution in Kenya") +
  theme(
    axis.text.x = element_text(size = 12, family = "Times", hjust = 1),
    axis.text.y = element_text(size = 12, family = "Times"),
    plot.title = element_text(size = 18, family = "Times"),
    axis.title = element_text(size = 14, family = "Times")
  ) 

```

## **Farming production**

```{r fig.width=12, fig.height=5}
#Farming production in the 5 biggest and smallest counties by production
smallest_producers <- merged_dataset |> 
  arrange(Farming) |> 
  slice_head(n = 5) |> 
  ggplot(aes(x = County, y = Farming)) +
  geom_col() +
  theme_classic() +
  labs(title = "Top 5 Counties with Smallest Farming Values",
       x = "",
       y = "Farming") + 
  scale_y_continuous(labels = comma, limits = c(0,400000)) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) 

biggest_producers <- merged_dataset |> 
  filter(County != "Kenya") |> 
  arrange(desc(Farming)) |> 
  slice_head(n = 5) |> 
  ggplot(aes(x = County, y = Farming)) +
  geom_col() +
  theme_classic() +
  labs(title = "Top 5 Counties with Biggest Farming Values",
       x = "",
       y = "") +
  scale_y_continuous(labels = comma, limits = c(0,400000)) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )
  

grid.arrange(smallest_producers, biggest_producers, ncol = 2, bottom = textGrob("County", vjust = -0.5))
```

## **Population**

```{r fig.width=12, fig.height=5}
#Population in the 5 biggest and smalles counties
smallest_population <- merged_dataset |> 
  arrange(Population) |> 
  slice_head(n = 5) |> 
  ggplot(aes(x = County, y = Population)) +
  geom_col() +
  theme_classic() +
  labs(title = "Top 5 Counties with Smallest Population",
       x = "",
       y = "Population") + 
  scale_y_continuous(labels = comma, limits = c(0,4400000)) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
  )



biggest_population <- merged_dataset |>
  filter(County != "Kenya") |> 
  arrange(desc(Population)) |> 
  slice_head(n = 5) |> 
  ggplot(aes(x = County, y = Population)) +
  geom_col() +
  theme_classic() +
  labs(title = "Top 5 Counties with Biggest Population",
       x = "",
       y = "") + 
  scale_y_continuous(labels = comma, limits = c(0,4400000)) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

grid.arrange(smallest_population, biggest_population, ncol = 2, bottom = textGrob("County", vjust = -0.5))
```

# Descriptives

## **Means**

```{r descriptives 1: means}
merged_dataset %>%
  summarise(
    Population = mean(Population),
    Farming = mean(Farming),
    "N of households" = mean(NumberOfHouseholds)
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Mean") %>%
  mutate(
    Mean = round(Mean, 0)
  ) |> 
  gt() %>%
  tab_header(
    title = md("**Descriptive statistics**"),
    subtitle = "Kenya census 2019"
  ) %>%
  fmt_number(
    columns = vars(Variable)
  )
```

## **Standard deviation**

```{r descriptives 2: SD}
merged_dataset |> 
  summarise(
    Population = sd(Population),
    Farming = sd(Farming),
    "N of households" = sd(NumberOfHouseholds)
  ) |> 
  pivot_longer(everything(), names_to = "Variable", values_to = "SD") |> 
  mutate(
    SD = round(SD, 0)
  ) |>
  gt() |> 
  tab_header(
    title = md("**Descriptive statistics**"),
    subtitle = "Kenya census 2019"
  ) |> 
  fmt_number(
    columns = c(Variable)
  )
```

## **Outliers**

```{r Descriptives 3: outliers, fig.width=12, fig.height=6}
##Population
outlier_population <- merged_dataset |> 
  filter(County != "Kenya") |> 
  ggplot(aes(x = 1, y = Population)) +
  geom_boxplot() +
  geom_text(aes(label = ifelse(Population < quantile(Population, 0.25) - 1.5 * IQR(Population) | 
                                  Population > quantile(Population, 0.75) + 1.5 * IQR(Population), County, "")),
            size = 5) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Outliers in Population",
    x = "",
    y = "Population"
  ) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14)
        )

##Farming
outlier_farming <- merged_dataset |> 
  filter(County != "Kenya") |> 
  ggplot(aes(x = 1, y = Farming)) +
  geom_boxplot() +
  geom_text(aes(label = ifelse(Farming < quantile(Farming, 0.25) - 1.5 * IQR(Farming) | 
                                  Farming > quantile(Farming, 0.75) + 1.5 * IQR(Farming), County, "")),
            size = 5) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Outliers in Farming",
    x = "",
    y = "Farming"
  ) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14)
        )

##Households
outlier_household <- merged_dataset |>
  filter(County != "Kenya") |> 
  ggplot(aes(x = 1, y = NumberOfHouseholds)) +
  geom_boxplot() +
  geom_text(aes(label = ifelse(NumberOfHouseholds < quantile(NumberOfHouseholds, 0.25) - 1.5 * IQR(NumberOfHouseholds) | 
                                  NumberOfHouseholds > quantile(NumberOfHouseholds, 0.75) + 1.5 * IQR(NumberOfHouseholds), County, "")),
            size = 5) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Outliers in Number of Households",
    x = "",
    y = "Number of Households"
  ) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14)
        )

#Conclusion: Even though there are three outliers, I will not remove them because they represent a natural occurence, where people congregate in urban areas, and the outliers are the biggest cities in Kenya.

grid.arrange(outlier_household, outlier_farming, outlier_population, ncol = 3)
```

#### Conclusion: Even though there are three outliers, I will not remove them because they represent a natural occurence, where people congregate in urban areas, and the outliers are the biggest cities in Kenya.

## **Plant production distribution**

```{r Descriptives 4: plant production}
#Create color coding
plants <- c("Tea", "Coffee", "Avocado", "Citrus", "Mango", "Coconut", "Macadamia", "Cashew Nut")
plant_colors <- c(
  Tea = "darkgreen",
  Coffee = "saddlebrown",
  Avocado = "limegreen",
  Citrus = "yellow",
  Mango = "darkorange",
  Coconut = "#d0d1e6",
  Macadamia = "darkolivegreen",
  "Cashew Nut" = "goldenrod"
)

#Create plot
merged_dataset |> 
  filter(County == "Kenya") |> 
  pivot_longer(cols = plants, names_to = "Crop", values_to = "Production") |> 
  ggplot(aes(x = fct_reorder(Crop, Production), y = Production)) +
  geom_bar(stat = "identity", fill = plant_colors) +
  coord_flip() +
  theme_classic() +
  labs(
    title = "Plant production in Kenya",
    x = "Crop",
    y = "Production"
  )
```

# Hypotheses

## 1. **Linear regression**

-   H1: Does the *population* affect the *farming production*?

-   H2: Is there a positive correlation between *Tea* and *Coffee* production?

## 2. **Mediation**

-   H3: *Population* is a mediator between *Farming* and *Coffee*

### **Hypothesis** 1.a: Linear regression

```{r H1, linear regression model 1}
#Does the population affect the farming production?

#Delete Kenya
counties_data <- merged_dataset |> 
  filter(County != "Kenya")

#Check if they correlate
cor(counties_data$Population, counties_data$Farming)

#Create plot
counties_data |> 
  ggplot(aes(x = Population, y = Farming)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Population and Farming Production",
    x = "Population",
    y = "Farming"
  ) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)

#Assumption 1.1: Normality
shapiro.test(counties_data$Population)
counties_data |> 
  ggplot(aes(x = Population)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Population",
    x = "Population",
    y = "Count"
  ) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)


shapiro.test(counties_data$Farming)
counties_data |> 
  ggplot(aes(x = Farming)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Farming",
    x = "Farming",
    y = "Count"
  ) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)

##Nairobi significantly alters the normality

pop_farming <- lm(Farming ~ Population, data = counties_data)

#Assumption 1.2: Linearity
raintest(pop_farming)
##Shows nonlinearity

#Assumption 1.3: Homoscedasticity
bptest(pop_farming)
##Shows heteroscedasticity

#Assumption 1.4: Normality of residuals
shapiro.test(pop_farming$residuals)
##Normality of residuals met

#Summary of model
glance(pop_farming)
```
#### Overall the model does not work with Nairobi in it, so I will remove it and see if it works without it. There is no normality and no homoscedasticity.


### **Hypothesis** 1.b: Linear regression without Nairobi

```{r H1, linear regression model 2}
#Assumption 2.1: Normality
counties_data_no_capital <- counties_data |> 
  filter(County != "Nairobi City")
shapiro.test(counties_data_no_capital$Population)

counties_data |> 
  ggplot(aes(x = Population)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Population",
    x = "Population",
    y = "Count"
  ) +
  theme_classic() +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)

##Removing Nairobi does not result in a normal population distribution

pop_farming_no_capital <- lm(Farming ~ Population, data = counties_data_no_capital)
glance(pop_farming_no_capital)

#Assumption 2.2: Linearity
raintest(pop_farming_no_capital)
##Shows linearity

#Assumption 2.3: Homoscedasticity
bptest(pop_farming_no_capital)
##Still shows heteroscedasticity

#Assumption 2.4: Normality of residuals
shapiro.test(pop_farming_no_capital$residuals)
##Normality of residuals not met
```
#### This model is still not sufficient, but it massively improves explanatory or predictive power, as can be seen by the R2 value.


### **Hypothesis** 1.c: Robust regression without Nairobi

```{r H1, robust regression}
#Final model
##Using the data without Nairobi so linearity is met

#Box-Cox transformation
bc <- boxcox(Farming ~ Population, data = counties_data_no_capital)
lambda <- bc$x[which.max(bc$y)]
box_cox_model <- lmrob(((Farming^lambda-1)/lambda) ~ Population, data = counties_data_no_capital)

qqnorm(box_cox_model$residuals)

#Huber regression
box_cox_model <- lmrob(Farming ~ Population, data = counties_data_no_capital)
summary(box_cox_model)

#Assumption 3.1: Normality of residuals
shapiro.test(box_cox_model$residuals)
##Assumption not met

#Assumption 3.2: Linearity
raintest(box_cox_model)
#Assumption met

#Assumption 3.3: Homoscedasticity
bptest(box_cox_model)
#Assumption not met

#Furthermore, robust regression is not suitable for less than 100 observations
```

#### In conclusion, there are too many issues with the data to create a reliable regression between population and farming, especially because of the issue of nonlinearity.

### **Hypothesis** 2: Linear regression

```{r H2: transformation}
##H2: Is there a positive correlation between Tea and Coffee production?
cor(counties_data_no_capital$Tea, counties_data_no_capital$Coffee, use = "pairwise.complete.obs")

counties_data_no_capital |> 
  ggplot(aes(x = Tea, y = Coffee)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Tea and Coffee Production",
    x = "Tea",
    y = "Coffee"
  ) +
  theme_minimal()

#A square root transformation was done to make the data more normally distributed.
counties_data_no_capital_sqrt = counties_data_no_capital |> 
  mutate(sqrt_tea = sqrt(Tea),
         sqrt_coffee = sqrt(Coffee))



counties_data_no_capital_sqrt |> 
  pivot_longer(cols = c(Tea, sqrt_tea), names_to = "Variable", values_to = "Value") |> 
  mutate(Variable = factor(Variable, levels = c("Tea", "sqrt_tea"))) |> 
  ggplot(aes(x = Value, fill = Variable)) +
  geom_density(alpha = 0.5, color = "black", size = 1) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Distribution of Tea and Square root Transformation",
       x = "Value",
       y = "Density") +
  theme_classic() +
  theme(legend.position = "none")
```

```{r H2: model}
tea_model <- lm(sqrt_tea ~ sqrt_coffee, data = counties_data_no_capital_sqrt)

#Assumption 4.1: Normality
shapiro.test(counties_data_no_capital_sqrt$sqrt_tea)
##Normality met

#Assumption 4.2: Linearity
raintest(tea_model)
##Linearity met

#Assumption 4.3: Homoscedasticity
bptest(tea_model)
##Homoscedasticity met

#Assumption 4.4: Normality of residuals
shapiro.test(tea_model$residuals)
##Normality of residuals met

#All assumptions met
```


## Hypothesis 2: **Results**

```{r H2 tables: Coefficients}
apa_table <- apa.reg.table(tea_model)
stats.table <- as.data.frame(summary(tea_model)$coefficients)
CI <- confint(tea_model)
stats.table <- cbind(row.names(stats.table), stats.table, CI)
names(stats.table) <- c("Variable", "Estimate", "Std. Error", "t-value", "p", "CI_lower", "CI_upper")
nice_table(stats.table, title = "Model coefficients", highlight = .05, note = c("CI = Confidence Interval", "* p < 0.05"))
```

```{r H2 tables: Model fit}
model_fit <- broom::glance(tea_model)
model_fit <- model_fit[, !(names(model_fit) %in% c("AIC", "BIC", "logLik", "nobs"))]
names(model_fit) <- gsub("sigma", "SD", names(model_fit), ignore.case = TRUE)
model_fit = model_fit |> 
  mutate(
    SD = as.numeric(SD),
    SD = round(SD, digits = 3)
  )
fit_table <- cbind(row.names(model_fit), model_fit)
nice_table(model_fit, title = "Model fit", note = c("* p < 0.05", "SD = Standard Deviation"))
```
#### There is a statistically significant moderate positive correlation between Tea and Coffee production, however there is a very limited sample size, because few counties have data abouth both Tea and Coffee production. The large confidence interval aligns with this conclusion. Coffee production explains 20% of the variance in Tea production.


## **Hypothesis** 3: Mediation

```{r H3, Mediation analysis}
#H3: Population is a mediator between farming and Coffee

#Check if they correlate
cor(counties_data_no_capital$Population, counties_data_no_capital$Farming, use = "pairwise.complete.obs")
cor(counties_data_no_capital$Population, counties_data_no_capital$Coffee, use = "pairwise.complete.obs")
cor(counties_data_no_capital$Farming, counties_data_no_capital$Coffee, use = "pairwise.complete.obs")

#Model building
mediation_model <- '
  # Regression equations
  Population ~ a * Farming
  Coffee ~ b * Farming + c * Population

  # Indirect effect (a * b)
  indirect := a * b

  # Total effect (c + indirect)
  total := c + indirect
'

mediation_results <- sem(mediation_model, data = counties_data_no_capital)
semPaths(mediation_results, what = "stand", whatLabels = "stand", style = "ram", rotation = 2, intercepts = FALSE, residuals = FALSE, layout = "tree2", nCharNodes = 0, sizeMan = 18, nCharEdges = 0, edge.label.cex = 2)
summary(mediation_results, standardized = TRUE)

#Assumptions need to be met for each regression path, I did not have time for this.

#Table
summary_table <- tidy(mediation_results)
kable(summary_table, caption = "Standardized Coefficients from Mediation Model")
```
